---
title: "peabody_intro"
output: html_document
date: '2022-05-26'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to the R environment

Although R is not new, it's popularity has increased rapidly over the last 10 years or so (see here for some interesting data). It was originally created and developed by Ross Ihaka and Robert Gentleman during the 1990's with the first stable version released in 2000. Nowadays R is maintained by the R Development Core Team. So, why has R become so popular and why should you learn how to use it? Some reasons include:

-   Open source and freely available.
-   Multi-platform (Windows, Mac, Linux)
-   Extensive and coherent set of tools for statistical analysis.
-   Extensive and flexible graphical facility capable of producing publication quality figures.
-   R has an expanding set of freely available 'packages' to extend R's capabilities.
-   R has an extensive support network with numerous online and freely available documents.

All of the reasons above are great reasons to use R. However, in our opinion, the single biggest reason to use R is that it facilitates robust and reproducible research practices. In contrast to more traditional 'point and click' software, writing code to perform your analysis ensures you have a permanent and accurate record of all the methods you used (and decisions you made) whilst analyzing your data. You are then able to share this code (and your data) with other researchers / colleagues / journal reviewers who will be able to reproduce your analysis exactly. This is one of the tenets of open science. We will cover other topics to facilitate open science throughout this book, including creating reproducible reports and version control

### Installing R

To get up and running the first thing you need to do is install R. R is freely available for Windows, Mac and Linux operating systems from the Comprehensive R Archive Network (CRAN) website . For Windows and Mac users we suggest you download and install the pre-compiled binary versions.

<https://cran.r-project.org/>

### Installing R-Studio

You can use the base installation of R, but we will be using a popular Integrated Development Environment (IDE) called RStudio. RStudio can be thought of as an add-on to R which provides a more user-friendly interface, incorporating the R Console, a script editor and other useful functionality.

<https://www.rstudio.com/>

### R-Studio orientation

The R-Studio IDE has a lot of features and more are being added all the time, but we will start with the basics of the Console, Script window, Environment tab, Files tab, Plots tab, Packages tab, and Help tab.

More advanced users will notice the Global Options for setting up additional parameters under the Tools menu.

### Alternatives to R-Studio

There are a lot of other IDEs that have R plug-ins such as VS Code and Jupyter notebooks. These IDEs are more general and also have integrations for Python, Java, and any number of other programming languages. We will restrict ourselves to R-Studio as it is the most advanced IDE that specializes in R (although you can use it for Python too).

### Understanding packages

Packages are the backbone of R. You can think of a package as a collection of functions, data, and help files collated into a well defined standard structure which you can download and install in R. These packages can be downloaded from a variety of sources but the most popular are CRAN and GitHub. CRAN is the official repository and packages available there have passed a rigorous screening process. GitHub is a website that hosts git repositories for all sorts of software and projects (not just R). Often, cutting edge development versions of R packages are hosted on GitHub so if you need all the new bells and whistles then this may be an option. However, a potential downside of using the development version of an R package is that it might not be as stable as the version hosted on CRAN (it's in development!) and updating packages won't be automatic.

To install a package from CRAN you can use the `install.packages()` function or use the Install button under the Packages tab in R-Studio.

To load a package into the environment, you can use either `library()` or `require()` . There are occasionally reasons for choosing one of these over the other, but typically you can use them interchangeably.

You can also use a package function without loading it into the environment by typing the package name followed by two semi-colons and the function name. For example, if you want to read an Excel file you can either use `library(readxl)` followed by the `read_excel()` function or simply use `readxl::read_excel()` .

Again, there are circumstances where you want to load the package into the environment and others where you don't. It just depends on the situation.

Since R is open-source, you can always open a function to see how it works. However, this can get a bit complex, so let's just start by looking at the code for one of the Base functions `cor()` from the `stats` package. To do this, just type the function name without the `()`.

```{r}
cor
```

### R Studio Projects

These are a great way of keeping things organised is to use RStudio Projects. An RStudio Project keeps all of your R scripts, R markdown documents, R functions and data together in one place. The nice thing about RStudio Projects is that each project has its own directory, workspace, history and source documents so different analyses that you are working on are kept completely separate from each other. This means that you can have multiple instances of RStudio open at the same time (if that's your thing) or you can switch very easily between projects without fear of them interfering with each other.

### Working Directory

The working directory is the default location where R will look for files you want to load and where it will put any files you save. One of the great things about using RStudio Projects is that when you open a project it will automatically set your working directory to the appropriate location. You can check the file path of your working directory by looking at bar at the top of the Console pane. Or by using the command line.

```{r}
getwd()
```

### Names

*There are only two hard things in Computer Science: cache invalidation and naming things.*\
*- Phil Karlton*

First, avoid using spaces or special characters (i.e. \@£\$%\^&\*(:/) in file names by replacing them with underscores or even hyphens. There are a lot of technical reasons for this, so you're just going to have to trust use on this one.

If you are versioning your files with sequential numbers (i.e. file1, file2, file3 ...) and you have more than 9 files you should use 01, 02, 03 .. 10 as this will ensure the files are printed in the correct order (see what happens if you don't). If you have more than 99 files then use 001, 002, 003... etc.

If your file names include dates, use the ISO 8601 format YYYY-MM-DD (or YYYYMMDD) to ensure your files are listed in proper chronological order.

For variable names, the same rules generally apply. If you have a space in a variable name, R will automatically fill it with a period. If you absolutely have to have a space in a variable name, you will need to surround the name with quotes or the \` symbol to ensure that R properly reads the name.

### Documentation

Be sure to use a lot of comments in your code so you can understand what you did later or someone else can. In R, the way to create a comment is with the \# symbol.

```{r}
print("This code prints")
# print("This code does not")
```

### Style

R is case sensitive i.e. A is not the same as a and anova is not the same as Anova.

R has a lot of specific instances when you need to specifically use parentheses `()`, braces `{}`, or brackets `[]`. So, you need to pay attention to what you're doing.

Because R is object-oriented, you will find yourself pushing data into variables a lot. You can use either a single equal sign `=` or the carrot `<-`. The carrot is typically recommended because there are instances where you need to specifically use the single `=` rather than the typical double `==`.

```{r}
my_first_variable <- "This is my first string variable"
print(my_first_variable)

my_first_numeric_variable = 1
print(my_first_numeric_variable)
```

# R basics

The common mathematical notation are fairly intuitive

```{r}
# addition
2 + 2
```

```{r}
# subtraction
10 - 5
```

```{r}
# multiplication
3 * 3
```

```{r}
# division
20 / 2
```

```{r}
# square root
sqrt(25)
```

```{r}
# exponent
2 ^ 2
```

Other standard functions...

```{r}
# create a vector
my_vec <- c(2, 3, 1, 6, 4, 3, 3, 7)
```

```{r}
# mean 
mean(my_vec)  
```

```{r}
# standard deviation 
sd(my_vec)
```

```{r}
# variance
var(my_vec)   
```

```{r}
# number of elements 
length(my_vec) 
```

```{r}
# number of unique elements
length(unique(my_vec))
```

```{r}
# create a sequence of numbers
seq(1:10)
```

### Extracting Elements

To extract (also known as indexing or subscripting) one or more values (more generally known as elements) from a vector we use the bracket `[]` notation. The general approach is to name the object you wish to extract from, then a set of square brackets with an index of the element you wish to extract contained within the square brackets. This index can be a position or the result of a logical test.

```{r}
# let's look at our vector again
my_vec
```

```{r}
# extract the 3rd element 
my_vec[3]
```

```{r}
# extract several elements from the middle
my_vec[3:8]
```

### Logical index

Another really useful way to extract data from a vector is to use a logical expression as an index. For example, to extract all elements with a value greater than 4 in the vector my_vec

```{r}
my_vec[my_vec > 4]
```

Here, the logical expression is `my_vec > 4` and R will only extract those elements that satisfy this logical condition. So how does this actually work? If we look at the output of just the logical expression without the square brackets you can see that R returns a vector containing either TRUE or FALSE which correspond to whether the logical condition is satisfied for each element. In this case only the 4th and 8th elements return a TRUE as their value is greater than 4.

```{r}
my_vec > 4
```

### Missing data

In R, missing data is usually represented by an `NA` symbol. Data may be missing for a whole bunch of reasons, maybe your machine broke down, maybe you broke down, maybe the weather was too bad to collect data on a particular day etc etc. Missing data can be a pain in the proverbial both from an R perspective and also a statistical perspective. From an R perspective missing data can be problematic as different functions deal with missing data in different ways. For example, let's say we collected air temperature readings over 10 days, but our thermometer broke on day 2 and again on day 9 so we have no data for those days

```{r}
temp  <- c(7.2, NA, 7.1, 6.9, 6.5, 5.8, 5.8, 5.5, NA, 5.5)
```

Now, if we try to calculate the mean with missing values...

```{r}
mean(temp)
```

We need to explicitly tell R to remove the missing values.

```{r}
mean(temp, na.rm = TRUE)
```

To find the missing values you can use the `is.na` for a Boolean response.

```{r}
is.na(temp)
```

You can also remove the `NA` values using the `na.omit()` function.

```{r}
mean(na.omit(temp))
```

### Reading and Writing data

There are several ways to read and write objects across a variety of different formats. Below are a few ways, but there are lots of different packages with different utilities and dependencies (i.e., Java).

```{r}

# Read delimited text files

read.table(file, header = FALSE, sep = "", quote = "\"'",
           dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
           row.names, col.names, as.is = !stringsAsFactors,
           na.strings = "NA", colClasses = NA, nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE,
           comment.char = "#",
           allowEscapes = FALSE, flush = FALSE,
           stringsAsFactors = FALSE,
           fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)


read.csv(file, header = TRUE, sep = ",", quote = "\"",
         dec = ".", fill = TRUE, comment.char = "", ...)


read.csv2(file, header = TRUE, sep = ";", quote = "\"",
          dec = ",", fill = TRUE, comment.char = "", ...)


read.delim(file, header = TRUE, sep = "\t", quote = "\"",
           dec = ".", fill = TRUE, comment.char = "", ...)


read.delim2(file, header = TRUE, sep = "\t", quote = "\"",
            dec = ",", fill = TRUE, comment.char = "", ...)
```

```{r}

# Write delimited files

write.table(x, file = "", append = FALSE, quote = TRUE, sep = " ",
            eol = "\n", na = "NA", dec = ".", row.names = TRUE,
            col.names = TRUE, qmethod = c("escape", "double"),
            fileEncoding = "")


write.csv(...)


write.csv2(...)
```

```{r}

# Read an excel file

library(readxl)

#read_excel() calls excel_format() to determine if path is xls or xlsx, based on the file extension and the file itself, in that order. Use read_xls() and read_xlsx() directly if you know better and want to prevent such guessing.

read_excel(path, sheet = NULL, range = NULL, col_names = TRUE,
  col_types = NULL, na = "", trim_ws = TRUE, skip = 0,
  n_max = Inf, guess_max = min(1000, n_max),
  progress = readxl_progress(), .name_repair = "unique")


read_xls(path, sheet = NULL, range = NULL, col_names = TRUE,
  col_types = NULL, na = "", trim_ws = TRUE, skip = 0,
  n_max = Inf, guess_max = min(1000, n_max),
  progress = readxl_progress(), .name_repair = "unique")


read_xlsx(path, sheet = NULL, range = NULL, col_names = TRUE,
  col_types = NULL, na = "", trim_ws = TRUE, skip = 0,
  n_max = Inf, guess_max = min(1000, n_max),
  progress = readxl_progress(), .name_repair = "unique")
```

```{r}

# Write an Excel file
library(writexl)

write_xlsx(
  x,
  path = tempfile(fileext = ".xlsx"),
  col_names = TRUE,
  format_headers = TRUE,
  use_zip64 = FALSE
)


# write an Excel file with multiple sheets

require(openxlsx)

list_of_datasets <- list("Name of DataSheet1" = dataframe1, 
                         "Name of Datasheet2" = dataframe2)
write.xlsx(list_of_datasets, file = "writeXLSX2.xlsx")
```

```{r}

# Read and write SAS, SPSS, or Stata files

library(haven)

# SAS
read_sas("mtcars.sas7bdat")
write_sas(mtcars, "mtcars.sas7bdat")

# SPSS
read_sav("mtcars.sav")
write_sav(mtcars, "mtcars.sav")

# Stata
read_dta("mtcars.dta")
write_dta(mtcars, "mtcars.dta")
```

### Data Types

There are only a few different data types in R.

-   **Numeric** data are numbers that contain a decimal. Actually they can also be whole numbers but we'll gloss over that.

-   **Integers** are whole numbers (those numbers without a decimal point).

-   **Logical** data take on the value of either `TRUE` or `FALSE`. There's also another special type of logical called `NA` to represent missing values.

-   **Character** data are used to represent string values. You can think of character strings as something like a word (or multiple words). A special type of character string is a *factor*, which is a string but with additional attributes (like levels or an order). We'll cover factors later.

R is usually able to determine the data type automatically, but you can check using the `class()` function.

```{r}
num <- 2.2
class(num)

```

```{r}

char <- "hello"
class(char)


```

```{r}

logi <- TRUE
class(logi)

```

It can sometimes be useful to be able to change the class of a variable using the `as.[className]()` family of coercion functions, although you need to be careful when doing this as you might receive some unexpected results (see what happens below when we try to convert a character string to a numeric).

Let's coerce a number to a character.

```{r}
class(num)
```

```{r}
num_char <-  as.character(num)
num_char
```

```{r}
class(num_char)
```

Now let's character to a number.

```{r}
class(char)
```

```{r}
char_num <- as.numeric(char)
```

### Data Structures

We're not going to get into all of the different data structures and details, but we will discuss the more common ones. We've already discussed vectors, so we will turn our attention to lists and data frames.

Lists are very flexible data structures which are ideal for storing irregular or non-rectangular data. To create a list we can use the `list()` function.

```{r}
list_1 <- list(c("black", "yellow", "orange"),
               c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE),
               matrix(1:6, nrow = 3))

list_1
```

Elements of a list can be extracted by indexing the levels.

```{r}
list_1[1]
```

```{r}
list_1[2]
```

```{r}
list_1[3]
```

Data frames are probably the most common structure you will encounter and are powerful two-dimensional objects made up of rows and columns which looks superficially very similar to a matrix. However, whilst matrices are restricted to containing data all of the same type, data frames can contain a mixture of different types of data.

Typically, in a data frame each row corresponds to an individual observation and each column corresponds to a different measured or recorded variable. This setup may be familiar to those of you who use Microsoft Excel to manage and store your data.

Use `library(help = "datasets")` to access a listing of R's built-in datasets.

```{r}
iris
```

```{r}
mtcars
```
