---
title: "2022 Summer R Workshop (Middle Part)"
output: html_document
date: '2022-06-16'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data and Packages
### Installing and loading packages
   * There are tidyverse, data.table, and base solutions to nearly every problem.
       + About the tidyverse (https://www.tidyverse.org/)
       + About the data.table (https://rdatatable.gitlab.io/data.table/)
       + About base R (https://www.statmethods.net/about/learningcurve.html)
   * In this workshop, we are going to focus on tidyverse. 

```{r packagetidyverse, echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE}
# install packages
install.packages("tidyverse") 
```

```{r package tidyverse, eval= TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# Loading packages 
library("tidyverse")
```

## Data Manipulation  
### Look at Data
```{r preview}
# first 6 rows with column names
head(iris)
```

```{r dim}
# how many rows of the data frame
nrow(iris)
# how many columns of the data frame
ncol(iris)
# what's the dimension of the data frame: row * column
dim(iris)
```
### Select Data by Column
```{r select columns}
# select only the width columns
width.only <- select(iris, Sepal.Width, Petal.Width)
head(width.only)
```
### Select Data by Row
```{r select rows}
# select sepal length greater than 6. 
long.length <- filter(iris, Sepal.Length > 6)
head(long.length)

# select the first 10 rows 
first.ten <- slice(iris, 1:10)
first.ten
```

### Select Data both by Column and Row
```{r select both}
# select sepal length greater than 6. and only the width columns
bothsel <- iris %>%
             filter(Sepal.Length > 6) %>%
             select(Sepal.Width, Petal.Width)    
head(bothsel)
dim(bothsel)
```
### Create New Columns based on Old Columns
```{r create column}
# add a new column in iris called as Sepal.Length_mm
addediris <- iris %>%
              mutate(Sepal.Length_mm = Sepal.Length * 100)
head(addediris)
dim(addediris)
```
### Rename Variables
```{r rename}
# rename(data frame, new_name = old_name, new_name = old_name)
iris <- rename(iris, pet.len = Petal.Length, pet.wid = Petal.Width) 
head(iris)
```
### Recode Values
```{r recode1}
# recode factors recode(old column, old name = new name)
newcol1 <- dplyr::recode(iris$Species, "setosa"="SET", "versicolor"="VER", "virginica" = "VIR")
newcol1
```

```{r recode2}
# pasting text together
newcol2 <- paste(iris$Species, newcol1, sep = "_")
newcol2
```

```{r recode3}
# logic from other variables
newcol3 <- ifelse(iris$Sepal.Length > 6 | iris$pet.len > 6, "long", "standard")
newcol3

# other logic format
#iris$Sepal.Length > 6 & iris$Petal.Length > 6
#iris$Sepal.Length <= 6 | iris$Petal.Length != 6
```

```{r recode4}
# converting column types
newcol4 <- as.factor(newcol3)
newcol4

# other converting types
# as.numeric()
# as.character()
```

### Aggregate and Sort
```{r aggregate column}
# create a summary table by species
sumtab <- iris %>%
            group_by(Species) %>%
            summarize(mean_Sepal.Length = mean(Sepal.Length), total = n())
sumtab
```

```{r sort column}
# sort summary table
sortedsumtab <- sumtab %>%
                  arrange(desc(mean_Sepal.Length))
sortedsumtab
```
### Combine and Merge
* Combine Columns
```{r cbind}
combinedcol <- cbind(width.only, select(iris, Species))
head(combinedcol)
dim(combinedcol)
```
* Combine Rows
```{r rbind}
iris.first.half <- slice(iris, 1:75)
iris.second.half <- slice(iris, 76:150)
back.together <- rbind(iris.first.half, iris.second.half)

dim(iris.first.half)
dim(iris.second.half)
dim(back.together)
```
* Merge or Join by Key Variable
```{r join}
# create a new data frame with the key variable called as "id.code"
id.code <- seq(1:nrow(iris))
some.rand <- rnorm(1:nrow(iris), mean = 0, sd = 1)
new.df <- data.frame(id.code, some.rand)
head(new.df)

# create a key variable in iris
iris <- iris %>% 
          mutate(id.code = seq(1:nrow(iris))) 

iris.join <- full_join(iris, new.df, by="id.code")
head(iris.join)

## right_join
## left_join
## inner_join
## anti_join
```

## Long Format and Wide Format Transformation
### Long to Wide
```{r long to wide}
longdata <- data.frame(ID = 1:3,
    expand.grid(Name = c("Dora", "John", "Rob"), Year = 2012:2014),
    BMI = round(runif(9, 18, 35), 0)
)
longdata

widedata <- longdata %>% pivot_wider(
              names_from = "Year", 
              names_prefix = "yr",
              values_from = "BMI")

widedata
```
### Wide to Long
```{r wide to long}
longdataback <- widedata %>% pivot_longer(
               cols = starts_with("yr"),
               names_to = "Year",
               values_to = "BMI",)

longdataback
```

## Analysis  
### Basic Statistics
```{r basic statistics}
mean(iris$Sepal.Length, na.rm = TRUE)
sd(iris$Sepal.Length, na.rm = TRUE)
cor(iris$Sepal.Length, iris$Sepal.Width)
```
### Statistical Tests
* t-test
```{r t-test}
iristtest <- filter(iris, Species!="setosa")
t.test(iristtest$Sepal.Length~iristtest$Species)
```
* anova
```{r anova}
one.way <- aov(Sepal.Length ~ Species, data = iris)
summary(one.way)
```
* regression
```{r regression}
my.model <- lm(Sepal.Length ~ pet.len, data=iris)
summary(my.model)
```

### Plots
```{r plot, message=FALSE}
my.plot <- ggplot(iris, aes(Sepal.Length, pet.len))+
   geom_point(aes(colour=Species))+
   geom_smooth(method = lm)+
   theme_classic()+
   ggtitle("MY PLOT")
my.plot
```

```{r save plot, message=FALSE}
ggsave("my.plot.jpg",my.plot, height=4, width=4, dpi=500)
```

## Programming
### Basic Rationale about Programming
```{r basic programming}
# Basic function structure
square.it <- function(x) {
  square <- x * x
  return(square)
}

square.it(5)
```
### If-else Structure
```{r ifelse}
# if else statement structure
num = 1000
if (num < 50) { 
  eval <- "small"
} else {
  eval <- "big"
}
eval
```
### For Loops
```{r for loop basics}
# Basic For Loop
for (i in 1:5) {
  print(i)
}
```

```{r for loops}
# Put For Loop in a Function
## newvector is an NA vector with the length as nrow(iris) 
newvector <- rep(NA, nrow(iris))
newvector

comvec <- function(newvector){
  for(i in 1:length(newvector)){
    newvector[i] <- paste("Demo", i, sep = "_")
  }
  print(newvector)
}

## now newvector is vector with filled in content of the same length. 
comvec(newvector)
```
### Avoid For Loops for Efficiency
```{r efficiency}
newvector <- paste("Demo", seq(1:nrow(iris)), sep="_")
newvector

## apply
## lapply
## sapply
## tapply
```
